Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-13 paper/ms.tex
Further discussion updates

\subsection{Connection to Cosmological Simulations}\label{ssec:cosmo}
As discussed in Section~\ref{sec:intro}, several authors have examined the formation of abundance plane structure in cosmological simulations. Of most interest to us is the zoom Au~23 in \citet{2018MNRAS.474.3629G}. This galaxy, one of six considered in their work, exhibits a bimodality that extends beyond the inner disk. The interpretation given by the authors is of a ``shrinking'' gaseous disk. This is equivalent to saying that the outer disk becomes depleted of gas. This shrinking of the disk, which occurs at $t_{\textrm{lookback}}\sim6\,\Gyr$, is associated with a dip in the SFR at that radius and a decrement in the median \alphaFe{} of $\sim0.05\,\dex$ (their Figure~2), which shortly after recovers. This sequence of events is more extended than in our work, but it strongly resembles the scenario in Figure~\ref{fig:before_after_sfh}.

@@

It was recently elaborated by \citet{2024arXiv240511025S} that these models also argue for a star formation gap between the two accretion episodes \citep[see also][]{1996ASPC...92..307G,1998A&A...338..161F,2000A&A...358..671G,2015A&A...578A..87S,2020A&A...640A..81N}. This gap is starvation-driven and can last several \Gyr{}. The present work argues for a starburst-driven quiescence followed by a rapid rejuvenation, with the entire process taking less than $1\,\Gyr$ and the gap only lasting a few hundred \Myr{}. The physical origin and some details are different, but one can appreciate that the bimodality arises from the same physical process.\footnote{Compare Section~\ref{ssec:formqui} to the first key result in the Conclusions of \citet{2024arXiv240511025S}.}

\subsection{Observational Connections: High Redshift Quenching}\label{ssec:obshiz}
If the Milky Way underwent a starburst to quiescent to rejuvenation sequence, as we argue in this paper, then an an analogue at $z\sim2$ would be identified as a post-starburst galaxy (PSB). With abundance matching, we expect the Milky Way's total stellar mass to be $\sim10^{10.3}\,\Msun$ at $z\sim2$ \citep{2013ApJ...771L..35V}. A number of authors have explored PSBs and quiescent galaxies at slightly higher masses at $z\sim2$, with large advances in the post-JWST era.

Furthermore, \citet{2023arXiv231215012C} found that lower mass quiescent galaxies (towards $10^{10.3}\,\Msun$) tend to be younger and more disky, pointing to a merger driven scenario. \citet{2023arXiv231212207A} found that at even lower masses ($<10^{9.5}\,\Msun$) There is also evidence that the AGN quenching mechanism, which we believe to be responsible for the quenching in our system (Appendix~\ref{app:cause_qui}), is operating at these redshifts \citep[e.g.][and references therein]{2023arXiv230806317D,2024arXiv240417945P,2024arXiv240518685M,2024Natur.630...54B}.

Overall, the starburst-driven quenching mechanism proposed in this work is allowed by high-redshift observations.

\subsection{Direct Search for Quiescence}\label{ssec:obsqui}
Figure~\ref{fig:before_after_sfh_by_iron} indicates a very direct observational test of the mechanism proposed in this work: for disk stars at a given \FeH{}, there should be a gap of $\sim300\,\Myr$ in ages at $\sim8\,\Gyr$, though in the Milky Way the gap could be larger. With a survey of properly chosen stars, this gap could be directly measured with a modestly sized (few hundred) sample of old stars with age uncertainties of $\sim1\%$. To our knowledge, the best method at these ages is differential analysis of solar twins, which can provide an age uncertainty of $\sim5\%$ \citep[e.g.][]{2014ApJ...795...23B,2018MNRAS.474.2580S}. However, this has only been applied to stars with solar metallicity, where there is not a clear separation between the high- and low-$\alpha$ sequences (though a gap in ages may still be present). A sort of differential approach could be applied also at lower metallicities, which would just lack the absolute age calibration that the Sun provides.

The gap could also be indirectly probed by a larger sample of slightly less precise ages. Astroseismology appears to be the most promising avenue. Currently the largest sample is APOKASC2 \citep{2018ApJS..239...32P}, which has $\sim3400$ stars with ages measured to $<20\%$ precision.\footnote{Errors here taken to be the maximum of the upper and lower age estimates in the public APOKASC2 catalog.} The upcoming PLATO mission is looking to measure $\sim20\textrm{k}$ stars with ages measured to $<10\%$ precision \citep{2024arXiv240605447R}. Further work is needed to determine how well these surveys could constrain the gap.

In addition to the presence of the gap, our work also predicts that: (1) the timing of the gap is \FeH{} dependent, and that (2) stars which form immediately after the gap are $\alpha$-poor compared to stars which form slightly later (see Figure~\ref{fig:alpha_vs_tform}). However, both of these tests rely on the precise age estimates.

\subsection{Future Work}
A natural next step would be to extend the idealized simulations in this work to a wider range of orbits, galaxy properties, and feedback models. However, the unrealism of our setup, which aids interpretation, limits its applicability to the real universe. And so something along the lines of the genetic modification technique to explore various mergers as done in this work may be useful \citep{2016MNRAS.455..974R,2017MNRAS.465..547P}.

>>>>

\subsection{Connection to Cosmological Simulations}\label{ssec:cosmo}
As discussed in Section~\ref{sec:intro}, several authors have examined the formation of abundance plane structure in cosmological simulations. Of most interest to us is the zoom Au~23 in \citet{2018MNRAS.474.3629G}. This galaxy, one of six considered in their work, exhibits a bimodality that extends beyond the inner disk. The interpretation given by the authors is of a ``shrinking'' gaseous disk. This is equivalent to saying that the outer disk becomes depleted of gas. This shrinking of the disk, which occurs at $t_{\textrm{lookback}}\sim6\,\Gyr$, is associated with a dip in the SFR at that radius and a decrement in the median \alphaFe{} of $\sim0.05\,\dex$ (their Figure~2), which shortly after recovers. This sequence of events is more extended than in our work, but it resembles the scenario in Figure~\ref{fig:before_after}.

@@

It was recently elaborated by \citet{2024arXiv240511025S} that these models also argue for a star formation gap between the two accretion episodes \citep[see also][]{1996ASPC...92..307G,1998A&A...338..161F,2000A&A...358..671G,2015A&A...578A..87S,2020A&A...640A..81N}. This gap is starvation-driven and can last several \Gyr{}. The present work argues for a starburst-driven quiescence followed by a rapid rejuvenation, with the entire process taking less than $1\,\Gyr$ and the gap only lasting a few hundred \Myr{}. The physical origin and some details are different, but one can appreciate that the bimodality arises from a similar process.\footnote{Compare Section~\ref{ssec:formqui} to the first key result in the Conclusions of \citet{2024arXiv240511025S}.}

\subsection{Direct Search for Quiescence}\label{ssec:obsqui}
Figure~\ref{fig:before_after_sfh_by_iron} indicates a very direct observational test of the mechanism proposed in this work: for disk stars at a given \FeH{}, there should be a gap of $\sim300\,\Myr$ in ages at $\sim8\,\Gyr$, though in the Milky Way the gap could be larger. With a survey of properly chosen stars, this gap could be directly measured with a modestly sized (few hundred) sample of old stars with age uncertainties of \edit1{a few percent}. To our knowledge, the best method at these ages is differential analysis of solar twins, which can provide an age uncertainty of $\sim5\%$ \citep[e.g.][]{2014ApJ...795...23B,2018MNRAS.474.2580S}. However, this has only been applied to stars with solar metallicity, where there is not a clear separation between the high- and low-$\alpha$ sequences (though a gap in ages may still be present). A sort of differential approach could be applied also at lower metallicities, which would simply lack the absolute age calibration that the Sun provides.

The gap could also be indirectly probed by a larger sample of slightly less precise ages. Astroseismology appears to be the most promising avenue. \edit1{Currently the largest sample is APOKASC-3 \citep{2025ApJS..276...69P}, which has $\sim2\textrm{k}$ stars with ages measured to $<12.5\%$ precision.\footnote{Errors here taken to be the maximum of the upper and lower age estimates in the APOKASC-3 catalog.}} The upcoming PLATO mission is looking to measure $\sim20\textrm{k}$ stars with ages measured to $<10\%$ precision \citep{2024arXiv240605447R}. Further work is needed to determine how well these surveys could constrain the gap.

\subsection{Future Work}\label{ssec:future_work}
\edit1{The largest unanswered question in the current work is the origin of gaps in the metallicity-dependent SFR. In Appendix~\ref{app:cause_qui}, we show that the merger is associated with strong feedback from the central AGN. Determining the precise mechanism and relationship between the two is delayed to future work. Furthermore, other mechanisms of quenching might be able to produce the gaps studied here, and so their exploration is worthwhile.}

\edit1{Another natural next step would be to extend the idealized simulations in this work to a wider range of orbits, galaxy properties, and feedback models. However, several aspects of our setup are unrealistic, for example, (1) the simulation is not in an expanding universe, (2) our feedback model is weaker than typical ones calibrated to full cosmological box simulations, (3) the initial conditions have a steeper potential well prior to star formation than in the real universe, and (4) our halos lack small-scale substructure. The simplicity of our setup aids interpretation, but also limits it applicability to the real unvierse. Something along the lines of the genetic modification technique to explore various mergers as done in this work may be useful \citep{2016MNRAS.455..974R,2017MNRAS.465..547P}.}
